---
title: "mwmap: A Complete Guide to Mapping Malawi in R"
subtitle: "Creating Publication-Ready Maps with Administrative Boundary Data"
author: "Chifundo Bita"
date: today
date-format: "MMMM D, YYYY"

format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: true
    number-depth: 3
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    fig-align: center
    fig-width: 7
    fig-height: 5
    df-print: kable
    smooth-scroll: true
    anchor-sections: true
    embed-resources: true
    css: |
      body {
        font-family: 'Source Sans Pro', Georgia, sans-serif;
        font-size: 16px;
        line-height: 1.8;
        color: #2c2c2c;
      }
      p { margin-bottom: 1.1em; }
      .callout { margin: 1.5em 0; }
      .callout-note    { border-left-color: #1a5a9a; }
      .callout-tip     { border-left-color: #238443; }
      .callout-warning { border-left-color: #e6550d; }
      h1, h2, h3, h4 { color: #1a3a5c; font-weight: 600; margin-top: 1.8em; }
      h1 { font-size: 2em; border-bottom: 3px solid #1a5a9a; padding-bottom: 0.3em; }
      h2 { font-size: 1.5em; border-bottom: 1px solid #dce6f0; padding-bottom: 0.2em; }
      code { color: #c0392b; background: #fdf6f6; padding: 1px 4px; border-radius: 3px; }
      pre  { background: #f8f8f8; border-left: 4px solid #1a5a9a; border-radius: 4px; }
      table { font-size: 0.93em; }
      thead { background: #eaf2fb; }
      .lead {
        font-size: 1.12em;
        color: #34495e;
        line-height: 1.9;
        margin-bottom: 1.4em;
      }

execute:
  echo: true
  warning: false
  message: false
  eval: true

knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
---

```{r}
#| label: setup
#| include: false
library(mwmap)
library(ggplot2)
library(dplyr)
library(readr)
```

# Introduction {.unnumbered}

::: {.lead}
**mwmap** is an R package that makes it easy to create accurate, professional
maps of Malawi using nothing more than a few lines of R code. It bundles
pre-processed spatial boundary data for every administrative level in the
country — from the national border down to individual Traditional Authorities —
and wraps that data in a family of functions that handle the tedious parts of
GIS work for you: joining your spreadsheet data to map geometries, normalising
inconsistent district name spellings, choosing colour scales, and producing
figures that are ready to drop straight into a report or publication.
:::

The package is built on top of **ggplot2** and **sf**, the two most widely used
R packages for data visualisation and spatial analysis respectively. If you
already know ggplot2, you will feel immediately at home — every mwmap function
returns a ggplot2 object, so you can add layers, themes, and scales using the
familiar `+` operator.

You do not need to be a GIS expert to use mwmap. You do not need to download
shapefiles, reproject coordinate systems, or understand spatial data formats.
All of that is handled internally. What you bring is your district-level data —
disease counts, survey estimates, economic indicators — and mwmap handles the
cartographic heavy lifting.

::: {.callout-note}
## Installation

**mwmap** is available on CRAN and can be installed in the usual way:

```r
# Stable release from CRAN
install.packages("mwmap")

# Latest development version from GitHub
# install.packages("remotes")
remotes::install_github("bitacanalytics/mwmap")
```

Once installed, load it alongside ggplot2:

```r
library(mwmap)
library(ggplot2)
```
:::

## What is in the package?

mwmap is organised into five functional groups. Understanding this structure
helps you find the right function quickly.

**Core mapping functions** produce the base map canvas. You will almost always
start with one of these before adding any data layers on top.

**Data and joining functions** are the bridge between your spreadsheet and the
map geometry. They handle the spatial join, clean up district names, and report
on how well your data matched the shapefile.

**Layer functions** add elements on top of an existing base map — highlighted
districts, point locations, text labels. These are designed to be chained
together using `+`.

**Styling functions** control colour scales and overall layout. The built-in
palettes have been chosen to work well for the most common Malawi mapping
use cases (health, agriculture, poverty, administration).

**Utility functions** are helpers you will use indirectly but that are also
useful on their own — particularly for cleaning district name columns before a
join.

| Category | Functions |
|---|---|
| **Core mapping** | `mw_map()`, `mw_regions()`, `mw_ta_map()` |
| **Data & joining** | `mw_join()`, `mw_choropleth()`, `mw_districts()`, `mw_districts_simple()` |
| **Map layers** | `mw_highlight()`, `mw_highlight_multi()`, `mw_points()`, `mw_labels()`, `mw_label_repel()` |
| **Styling** | `mw_layout()`, `scale_fill_mw()`, `get_mw_palette()`, `get_mw_palettes()`, `view_mw_palettes()` |
| **Utilities** | `mw_clean_names()`, `mw_suggest_matches()` |

: mwmap function groups {tbl-colwidths="[30,70]"}

The package also ships five **built-in spatial datasets**. These are `sf`
objects — essentially data frames where one column contains polygon geometry —
sourced from the Malawi Spatial Data Platform (MASDAP) and pre-processed for
use in R.

| Dataset | Description | Polygons |
|---|---|---|
| `mw_level_0` | National boundary of Malawi | 1 |
| `mw_level_1` | The three administrative regions | 3 |
| `mw_level_2` | All 28 districts | 28 |
| `mw_level_3` | Traditional Authorities (sub-district) | ~250 |
| `major_lakes` | Lake Malawi, Lake Malombe, Lake Chilwa | 3 |

: Spatial datasets bundled in mwmap {tbl-colwidths="[22,55,23]"}

You can use these datasets directly in your own `sf` or `ggplot2` code, but in
most cases the mwmap functions will load them automatically for you.

---

# Getting Started

## Your first map

The fastest way to get started is to call `mw_map()` with no arguments. This
produces a clean, grey district-level outline of Malawi in a single line of
code.

```{r}
#| label: fig-first-map
#| fig-cap: "A minimal Malawi map produced by mw_map() with no arguments."
library(mwmap)
library(ggplot2)

mw_map()
```

What you get back is a standard **ggplot2 object** — the same kind of object
you get from calling `ggplot()`. This is important because it means you can
immediately extend the map using any ggplot2 syntax you already know. Want to
add a title? Use `labs()`. Want to change the background colour? Use `theme()`.
Want to add a second layer of points or polygons? Just use `+`.

```{r}
#| label: fig-first-map-titled
#| fig-cap: "The same map with a title and caption added using standard ggplot2 labs()."
mw_map() +
  labs(
    title = "Malawi Administrative Districts",
    caption = "Source: MASDAP / Department of Surveys"
  )
```

## Understanding administrative levels

Malawi's administrative geography has four nested levels. mwmap bundles data
for all four, and you switch between them using the `level` argument in
`mw_map()`.

- **Level 0** is the national boundary — useful as a reference outline or
  backdrop behind other data.
- **Level 1** shows the three regions: Northern, Central, and Southern.
  Regional comparisons — vaccination coverage, food security — are a natural
  fit for this level.
- **Level 2** is the default and the most commonly used level. Malawi's 28
  districts are the primary unit for government statistics, health surveillance,
  and most national surveys.
- **Level 3** shows Traditional Authorities, the sub-district units governed
  by traditional leaders. This is the finest grain available in the package and
  is useful for local-level planning and community health work.

```{r}
#| label: fig-levels
#| fig-cap: "All four administrative levels available in mwmap. Level 2 (districts) is the default."
#| layout-ncol: 2
mw_map(level = 0) + labs(title = "Level 0 — National boundary")
mw_map(level = 1) + labs(title = "Level 1 — Three regions")
mw_map(level = 2) + labs(title = "Level 2 — 28 districts (default)")
mw_map(level = 3) + labs(title = "Level 3 — Traditional Authorities")
```

Notice how level 3 is considerably more detailed. Because Traditional Authority
boundaries are complex, rendering the full level-3 map is slower than the other
levels. If you only need a subset of TAs — for example, those in a single
district — use `mw_ta_map(districts = "Lilongwe")` rather than
`mw_map(level = 3)` to keep rendering fast.

## Filtering to a region or set of districts

You will often want to focus on a subset of the country rather than mapping
everything at once. mwmap gives you two built-in filters:

- `region` accepts the full region name (`"Northern"`, `"Central"`,
  `"Southern"`) or a single-letter shortcut (`"n"`, `"c"`, `"s"`).
- `districts` accepts a character vector of district names. Matching is
  case-insensitive.

When you filter to a region, `mw_map()` automatically crops the bounding box
so the selected area fills the plot. You do not need to adjust any axis limits
manually.

```{r}
#| label: fig-filter
#| fig-cap: "Filtering to a single region (left) or a custom set of districts (right)."
#| layout-ncol: 2

# Shortcut "s" is equivalent to "Southern"
mw_map(region = "s") +
  labs(title = "Southern Region")

# Show Lilongwe and its neighbouring districts
mw_map(districts = c("Lilongwe", "Dowa", "Dedza", "Ntcheu", "Mchinji")) +
  labs(title = "Lilongwe and Neighbouring Districts")
```

---

# The Core Mapping Functions

## `mw_map()` — the main workhorse {#sec-mw-map}

`mw_map()` is the central function in the package. Nearly every workflow starts
here. It is designed to be useful at both extremes — a single call with no
arguments produces a clean outline, while a heavily parameterised call can
produce a complete choropleth with highlights, labels, and lake overlay in one
step.

The table below covers the most important arguments. A complete reference is
available via `?mw_map`.

| Argument | Default | What it does |
|---|---|---|
| `level` | `2` | Administrative level: 0 = country, 1 = regions, 2 = districts, 3 = TAs |
| `data` | `NULL` | A data frame or sf object with values to display |
| `fill` | `NULL` | Column name (as a string) to use for choropleth fill colour |
| `fill_color` | `"#f0f0f0"` | Solid fill colour when no `fill` column is provided |
| `lakes` | `FALSE` | Whether to overlay the three major lakes |
| `lake_color` | `"#9ecae1"` | Fill colour for lakes |
| `borders` | `TRUE` | Whether to draw polygon borders |
| `border_color` | `"#333333"` | Border colour |
| `border_size` | `0.3` | Border line width |
| `region` | `NULL` | Restrict map to one region |
| `districts` | `NULL` | Restrict map to named districts |
| `label_districts` | `FALSE` | Add district name labels at centroids |
| `label_size` | `3` | Size of district labels |
| `highlight_districts` | `NULL` | Districts to draw with a bold coloured border |
| `highlight_border_color` | `"#E41A1C"` | Colour of the highlight border |
| `highlight_fill` | `NA` | Fill colour for highlighted districts (NA = transparent) |
| `interactive` | `FALSE` | Return a plotly interactive map instead of ggplot2 |
| `title`, `subtitle`, `caption` | `NULL` | Map text elements |

: Key `mw_map()` arguments {tbl-colwidths="[28,18,54]"}

### Adding the major lakes

Malawi is defined by water as much as by its land. Lake Malawi covers roughly
a fifth of the country's total area, and Lakes Malombe and Chilwa are
significant features in the south. For most maps, adding the lakes makes the
geography immediately recognisable to a Malawian audience and gives the reader
an important spatial reference point.

Set `lakes = TRUE` to overlay them. You can adjust the fill colour with
`lake_color` and the border with `lake_border_color`.

```{r}
#| label: fig-lakes
#| fig-cap: "District map with major lakes overlaid. The lakes are stored as a separate sf dataset (major_lakes) and drawn on top of the district polygons."
mw_map(
  lakes = TRUE,
  lake_color = "#6BAED6",
  lake_border_color = "#3182BD"
) +
  labs(title = "Malawi — Districts and Major Lakes")
```

### Highlighting specific districts

`mw_map()` has built-in support for drawing attention to specific districts
without needing a separate `mw_highlight()` layer. Use `highlight_districts`
to name the districts you want to emphasise.

By default, the highlight draws a bold red border around the named districts.
You can customise the border colour, border width, and fill colour. Setting
`highlight_fill = NA` (the default) makes the fill transparent — only the bold
border draws attention while the underlying fill colour remains visible.

```{r}
#| label: fig-highlight-inline
#| fig-cap: "Three major urban districts highlighted with a bold coloured border and a light orange fill."
mw_map(
  highlight_districts = c("Lilongwe", "Blantyre", "Mzimba"),
  highlight_border_color = "#E41A1C",
  highlight_border_size = 1.5,
  highlight_fill = "#FEEDDE"
) +
  labs(
    title = "Major Urban Districts",
    caption = "Highlighted: Lilongwe, Blantyre, Mzimba"
  )
```

### Adding district name labels

Set `label_districts = TRUE` to add each district name at the centroid of its
polygon. Labels that would overlap are automatically suppressed. For finer
control — labelling only specific districts, changing the font, or using
ggrepel for non-overlapping placement — use the standalone `mw_labels()`
function described in @sec-labels.

```{r}
#| label: fig-labels-inline
#| fig-cap: "All 28 district names added via label_districts = TRUE."
mw_map(
  fill_color = "#f7f7f7",
  border_color = "grey60",
  label_districts = TRUE,
  label_size = 2.5,
  label_color = "#333333"
) +
  labs(title = "Malawi District Names")
```

---

## `mw_regions()` — regional overview {#sec-mw-regions}

`mw_regions()` is a convenience function that aggregates the 28 district
polygons into the three administrative regions and plots them. Internally it
calls `dplyr::group_by()` and `sf::st_union()` to dissolve district boundaries,
so you do not need to pre-process the data yourself.

Use this function when your analysis is at the regional level — comparing
Northern, Central, and Southern on some indicator — or when you need a clean
regional overview to orient the reader at the start of a report.

The `region_colors` argument accepts a named vector where the names are
`"Northern"`, `"Central"`, and `"Southern"`. This makes it straightforward to
encode data values as fill colours by pre-computing the colour for each region
and passing the result in.

```{r}
#| label: fig-regions
#| fig-cap: "The three administrative regions with distinct colours, name labels, and the major lakes overlaid."
mw_regions(
  region_colors = c(
    "Northern" = "#E41A1C",
    "Central" = "#377EB8",
    "Southern" = "#4DAF4A"
  ),
  labels = TRUE,
  label_size = 5,
  label_color = "white",
  label_face = "bold",
  lakes = TRUE
) +
  labs(title = "Administrative Regions of Malawi")
```

Notice that `labels = TRUE` places each region name at the geometric centroid
of the dissolved polygon. For the Southern Region — which wraps around Lake
Malawi — the centroid may not look perfectly centred visually, but it is the
mathematical centre of the polygon.

---

## `mw_ta_map()` — traditional authorities {#sec-mw-ta-map}

Traditional Authorities (TAs) are the sub-district administrative units in
Malawi, each governed by a customary chief. The package includes boundaries for
approximately 250 TAs sourced from MASDAP. `mw_ta_map()` is a dedicated
function for rendering these boundaries, with several features designed for
working at this fine scale.

The most important argument is `districts`. Rendering all 250 TAs at once is
slow and the resulting map is very dense. It is almost always better to filter
to the district or districts you care about. When you do, `mw_ta_map()` also
draws the district boundary on top of the TA polygons by default
(`district_border = TRUE`), which helps the reader understand where the
selected TAs sit within the broader geography.

```{r}
#| label: fig-ta-all
#| fig-cap: "All Traditional Authorities. Thin grey lines are TA boundaries; thicker red lines are district boundaries drawn on top."
mw_ta_map(
  border_color = "grey70",
  border_size = 0.15,
  district_border = TRUE,
  district_border_color = "#E41A1C",
  district_border_size = 0.6
) +
  labs(
    title = "Traditional Authorities of Malawi",
    caption = "Thin grey = TA boundaries · Thick red = district boundaries"
  )
```

Filtering to specific districts dramatically improves readability and speed.

```{r}
#| label: fig-ta-filter
#| fig-cap: "TAs in just two districts. Filtering with the districts argument makes the map much more readable and faster to render."
mw_ta_map(
  districts = c("Lilongwe", "Dowa"),
  labels = TRUE,
  label_size = 2,
  label_color = "#333333",
  district_border_color = "black",
  district_border_size = 0.9
) +
  labs(title = "Traditional Authorities — Lilongwe and Dowa")
```

::: {.callout-tip}
## `mw_ta_map()` vs `mw_map(level = 3)`

Both produce TA-level maps but are designed for different situations.
Use `mw_map(level = 3)` when you want to layer TA boundaries on top of
other `mw_map()` features like fill data, highlights, or points. Use
`mw_ta_map()` when you want the dedicated district-border overlay and the
`districts` filter, which are specific to this function.
:::

---

# Choropleth Mapping

A choropleth map uses colour to encode a quantitative or categorical variable,
with each polygon coloured according to its value. This is by far the most
common type of thematic map in research and reporting — malaria incidence by
district, poverty rates by region, vaccination coverage by TA.

mwmap gives you two workflows for choropleth mapping: a **quick one-step
route** via `mw_choropleth()`, and a **flexible two-step route** via
`mw_join()` + `mw_map()`. Both produce the same kind of output; the choice
depends on how much control you need over the final appearance.

## Quick route — `mw_choropleth()` {#sec-choropleth}

`mw_choropleth()` is designed for speed. You pass it a data frame, the name of
the column you want to map (unquoted, like a tidyverse verb), and optionally a
palette name. It handles the join, the scale, and the plot in a single call.

The key arguments are:

- `data` — your data frame. It must contain a column of district names.
- `value` — the column to map, passed **unquoted** (e.g. `malaria_cases`,
  not `"malaria_cases"`).
- `district_col` — the name of the district name column in your data. Defaults
  to `"district"`.
- `palette` — one of the 25 named mwmap palettes. Defaults to `"health"`.
- `scale_type` — `"auto"` (the default) detects whether your variable is
  numeric (continuous scale) or a factor (discrete scale).
- `lakes` — whether to overlay major lakes.
- `title`, `subtitle`, `caption` — text elements for the map.

To demonstrate, we will use a simulated dataset of malaria indicators for all
28 districts. This same dataset is reused throughout the rest of this guide.

```{r}
#| label: health-data
# Reading simulated district-level health data for all 28 districts

health_data <- read_csv("Malaria_data.csv")

head(health_data)

```

```{r}
#| label: fig-choropleth-basic
#| fig-cap: "A basic choropleth of malaria case counts. Note how Lilongwe and Blantyre dominate simply because they have the largest populations — not necessarily the highest risk."
mw_choropleth(
  health_data,
  malaria_cases,
  palette = "malaria",
  title = "Reported Malaria Cases by District",
  subtitle = "Simulated data for illustration",
  caption = "Source: Simulated · mwmap package"
)
```

### Why you should map rates, not raw counts

The map above shows Lilongwe and Blantyre with the highest case counts. But
those are also Malawi's two most populous districts — of course they have more
cases. This is a classic pitfall in choropleth mapping. A map of raw counts
often tells you more about population distribution than about the phenomenon
you are trying to understand.

The solution is to compute a **rate** — dividing the count by the population
to put all districts on a comparable footing. A rate per 1,000 population is a
standard epidemiological measure.

```{r}
#| label: fig-choropleth-rate
#| fig-cap: "The same data expressed as a rate per 1,000 population. The geographic pattern is now very different — smaller, more rural districts may show higher burden than the urban centres."
health_data$rate_per_1000 <- round(
  health_data$malaria_cases / health_data$population * 1000,
  1
)

mw_choropleth(
  health_data,
  rate_per_1000,
  palette = "malaria",
  title = "Malaria Incidence Rate by District",
  subtitle = "Cases per 1,000 population · Simulated data",
  legend_title = "Rate per 1,000",
  caption = "Source: Simulated · mwmap package",
  n_breaks = 5
)
```

Comparing these two maps is instructive. The count map is dominated by the two
big cities; the rate map may look completely different, revealing high-burden
districts that were invisible in the count map.

### Categorical (discrete) choropleths

Not all variables are continuous numbers. Sometimes you want to map a category —
risk tier, intervention group, funding status. When your variable is a
**factor**, `mw_choropleth()` detects this automatically and switches to a
discrete colour scale. We use `cut()` with `ordered_result = TRUE` to create
an ordered factor, which tells the colour scale to arrange the legend from
low to high.

```{r}
#| label: fig-choropleth-cat
#| fig-cap: "A categorical choropleth. mw_choropleth() detects the factor variable automatically and uses a discrete scale."
health_data$risk_tier <- cut(
  health_data$rate_per_1000,
  breaks = c(0, 4, 7, 10, 15, Inf),
  labels = c("Very Low", "Low", "Medium", "High", "Very High"),
  right = TRUE
)

mw_choropleth(
  health_data,
  risk_tier,
  palette = "health",
  title = "Malaria Risk Tier by District",
  legend_title = "Risk Level",
  caption = "Source: Simulated · mwmap package"
)
```

---

## Flexible route — `mw_join()` + `mw_map()` {#sec-join}

When you need more control than `mw_choropleth()` offers — a custom scale,
additional layers, multiple fills — the two-step approach gives you full
flexibility. The idea is simple:

1. **Join** your data to the map geometry using `mw_join()`, producing an `sf`
   object where every district row is enriched with your data columns.
2. **Plot** that joined `sf` object using `mw_map(data = ..., fill = ...)`.

Because the result of step 1 is an `sf` object, you can then add any ggplot2
scales, `mw_highlight()` layers, `mw_points()` layers, or custom themes.

```{r}
#| label: fig-join-workflow
#| fig-cap: "The two-step mw_join() + mw_map() workflow. Joining first gives you full control over scales and additional layers."

# Step 1: join your data frame to the district polygons
joined <- mw_join(
  data = health_data,
  district_col = "district",
  quiet = TRUE
)

# Step 2: pass the joined sf object to mw_map()
mw_map(
  data = joined,
  fill = "rate_per_1000",
  lakes = TRUE,
  border_color = "white",
  border_size = 0.4,
  title = "Malaria Incidence Rate",
  subtitle = "Cases per 1,000 population"
) +
  scale_fill_mw(
    palette = "malaria",
    name = "Rate per 1,000",
    na.value = "grey80"
  )
```

### How `mw_join()` works

`mw_join()` is more than a simple merge. Under the hood it does several things
that make it far more robust than calling `dplyr::left_join()` directly:

**Name standardisation.** Before attempting to match, `mw_join()` passes all
district names through `mw_clean_names()` automatically. This normalises
capitalisation, removes trailing words like "District" or "TA", and fixes known
variants (for example, "Nkhata Bay" becomes "Nkhatabay" to match the shapefile).
Most of the time your data will join cleanly without any pre-processing.

**A matching report.** By default (`verbose = TRUE`), `mw_join()` prints a
summary showing how many of your districts matched the shapefile, which ones
did not match, and — for the unmatched ones — a list of suggested alternatives
based on string similarity. This is invaluable for catching typos early.

**A left join by default.** `mw_join()` keeps all 28 district rows from the
shapefile regardless of whether your data has a value for that district
(`keep_all = TRUE`). Districts with no data appear grey on the map. This
prevents the common mistake of accidentally hiding districts with missing data.

```{r}
#| label: join-report
# Run mw_join() without quiet = TRUE to see the matching report.
# This is especially useful when working with a new dataset for the first time.
joined_verbose <- mw_join(
  data.frame(
    district = c("lilongwe", "BLANTYRE", "Mzimba", "Zomba City"),
    value = c(120, 200, 85, 60)
  ),
  district_col = "district"
)
# mw_join() will tell you that "Zomba City" did not match and suggest "Zomba"
```

---

# Adding Map Layers

One of the strengths of the mwmap + ggplot2 approach is that maps are
**composable**. You build a base map and then stack layers on top using `+`.
This section covers the main layer functions.

## `mw_highlight()` — draw attention to districts {#sec-highlight}

`mw_highlight()` overlays a coloured fill or bold outline on specific districts.
It is useful whenever you want to direct the reader's eye to particular areas —
the districts covered by a programme, the ones that exceed a threshold, the
capital and major cities.

The key arguments are:

- `districts` — district names to highlight (case-insensitive).
- `fill` — the fill colour. Defaults to `"red"`.
- `color` — the border colour. Defaults to `"black"`.
- `alpha` — fill transparency, from 0 (invisible) to 1 (opaque).
- `size` — border line width.
- `outline_only` — if `TRUE`, suppresses the fill and draws only the border.
  Ideal for adding emphasis on top of a choropleth without obscuring data values.
- `label` — if `TRUE`, adds district name labels inside highlighted areas.

```{r}
#| label: fig-highlight
#| fig-cap: "Three lakeshore districts highlighted with a gold fill. The highlight layer sits on top of the base map and is styled independently."
mw_map(fill_color = "#f0f0f0", border_color = "grey70") +
  mw_highlight(
    districts = c("Salima", "Nkhotakota", "Mangochi"),
    fill = "#FEC44F",
    color = "#E6550D",
    alpha = 0.85,
    size = 0.8
  ) +
  labs(
    title = "Lakeshore Districts of Malawi",
    subtitle = "Salima · Nkhotakota · Mangochi"
  )
```

### Using `outline_only = TRUE` on a choropleth

When you have a choropleth showing continuous values, a solid fill highlight
would cover the data. The `outline_only` option draws just the border, adding
visual emphasis without obscuring the underlying colours.

```{r}
#| label: fig-highlight-outline
#| fig-cap: "A bold outline highlight layered on top of a choropleth. The fill values remain fully visible."
mw_choropleth(health_data, rate_per_1000, palette = "health", quiet = TRUE) +
  mw_highlight(
    districts = c("Lilongwe", "Blantyre"),
    outline_only = TRUE,
    color = "#E41A1C",
    size = 1.2
  ) +
  labs(caption = "Red outline = major urban centres")
```

## `mw_highlight_multi()` — a different colour per district

When you want to highlight several districts but give each one a distinct
colour — to show which programme operates in each area, or to indicate
different intervention types — `mw_highlight_multi()` accepts a **named
vector** where the names are district names and the values are colours.

```{r}
#| label: fig-highlight-multi
#| fig-cap: "Each highlighted district has its own colour, defined by a named character vector."
mw_map(fill_color = "#f7f7f7", border_color = "grey70") +
  mw_highlight_multi(
    c(
      "Lilongwe" = "#377EB8",
      "Blantyre" = "#E41A1C",
      "Mzuzu City" = "#4DAF4A",
      "Zomba" = "#FF7F00"
    ),
    alpha = 0.8
  ) +
  labs(title = "Major Cities of Malawi")
```

---

## `mw_points()` — plot geographic locations {#sec-points}

`mw_points()` adds a layer of points to a map from a data frame containing
longitude and latitude columns. Common uses include health facility locations,
survey cluster coordinates, school or market sites, and infrastructure.

The `lon` and `lat` arguments are passed **unquoted** — like a tidyverse verb.
The function converts the data frame to an `sf` object internally and uses
`geom_sf()` so the coordinate system integrates correctly with the base map.

Important: when mapping a column to `color` or `size`, pass the **column name
as a quoted string** — e.g. `color = "facility_type"`, not `color = facility_type`.
This is different from how `mw_choropleth()` works (which uses unquoted names),
because `mw_points()` needs to handle both fixed values (e.g. `color = "red"`)
and column names in the same argument.

```{r}
#| label: facilities-data
# Sample health facility data
facilities <- data.frame(
  name = c(
    "Kamuzu Central", "Queen Elizabeth Central", "Mzuzu Central","Zomba Central",
    "Dedza District", "Salima District", "Karonga District", "Mangochi District"),
  lon = c(33.780, 35.002, 34.018, 35.320, 34.339, 34.459, 33.930, 35.264),
  lat = c(-13.978, -15.785, -11.462, -15.380, -14.381, -13.782, -9.932, -14.477),
  level = c(
    "Central", "Central", "Central", "Central", "District", "District", "District", "District" ),
  beds = c(1200, 1350, 600, 450, 180, 160, 200, 210)
)
```

```{r}
#| label: fig-points-basic
#| fig-cap: "Health facilities plotted as red triangles on a district base map."
mw_map(fill_color = "#f5f5f5", border_color = "grey60", lakes = TRUE) +
  mw_points(
    facilities,
    lon,
    lat,
    color = "#E41A1C",
    size = 3,
    shape = 17 # 17 = filled triangle; 19 = filled circle
  ) +
  labs(
    title = "Central Referral and District Hospitals",
    caption = "Source: Ministry of Health, Malawi (simulated)"
  )
```

### Mapping point attributes to aesthetics

Pass a **quoted column name** to `color` or `size` to map a data attribute to
a visual property. Then add a standard ggplot2 `scale_color_manual()` or
`scale_size_continuous()` to control the exact colours and sizes used.

```{r}
#| label: fig-points-aes
#| fig-cap: "Facility level mapped to point colour. Central hospitals are shown in red, district hospitals in blue."
mw_map(fill_color = "#f0f4f8", border_color = "grey70", lakes = TRUE) +
  mw_points(
    facilities,
    lon,
    lat,
    color = "level", # quoted column name
    size = 3,
    alpha = 0.9
  ) +
  scale_color_manual(
    values = c("Central" = "#E41A1C", "District" = "#1a5a9a"),
    name = "Facility Level"
  ) +
  labs(
    title = "Health Facilities by Level",
    subtitle = "Central and district hospitals"
  )
```

### Labelling points

Set `label = TRUE` and provide the column name as a string to `label_column`
to add text labels next to each point. Labels are automatically suppressed when
they would overlap. For repelled, non-overlapping labels, use
`repel_labels = TRUE` (requires the `ggrepel` package).

```{r}
#| label: fig-points-labels
#| fig-cap: "Central hospitals with name labels attached at point coordinates."
hospitals <- facilities[facilities$level == "Central", ]

mw_map(fill_color = "#f5f5f5", border_color = "grey70") +
  mw_points(
    hospitals,
    lon,
    lat,
    color = "#1a5a9a",
    size = 3,
    label = TRUE,
    label_column = "name", # column name as a quoted string
    label_size = 2.8,
    label_color = "#333333"
  ) +
  labs(title = "Central Hospitals — Malawi")
```

---

## `mw_labels()` — district name labels {#sec-labels}

`mw_labels()` adds district name text at the centroid of each polygon. It is
the standalone version of the `label_districts = TRUE` argument in `mw_map()`
and gives you more control: you can restrict labels to a subset of districts,
change the font, and pass any additional arguments through to `geom_sf_text()`.

The most useful arguments are:

- `districts` — a character vector of district names to label. If `NULL`
  (default), all districts are labelled.
- `size` — text size. The default of 3 works well for a full-country map.
- `color` — text colour.
- `fontface` — `"plain"`, `"bold"`, `"italic"`, or `"bold.italic"`.
- `label_column` — column to use for the label text. Defaults to `"ADM2_EN"`.

```{r}
#| label: fig-labels
#| fig-cap: "All 28 districts labelled at their centroids."
mw_map(fill_color = "#EDF8E9", border_color = "grey60") +
  mw_labels(
    size = 2.5,
    color = "#1a3a5c",
    fontface = "bold"
  ) +
  labs(title = "Malawi — All Districts Labelled")
```

A common pattern is to use `mw_labels()` to label only the specific districts
you have highlighted, rather than all 28. This keeps the map clean while still
giving the reader the names they need.

```{r}
#| label: fig-labels-selected
#| fig-cap: "Labels only on the four highlighted cities. Restricting labels to highlighted districts avoids clutter."
mw_map(
  highlight_districts = c("Lilongwe", "Blantyre", "Mzuzu City", "Zomba"),
  highlight_fill = "#FEEDDE",
  highlight_border_color = "#E6550D",
  fill_color = "#f7f7f7",
  border_color = "grey70"
) +
  mw_labels(
    districts = c("Lilongwe", "Blantyre", "Mzuzu City", "Zomba"),
    size = 3.5,
    color = "#E6550D",
    fontface = "bold"
  ) +
  labs(title = "Major Cities of Malawi")
```

---

# Colour Palettes

Choosing colours for a choropleth map is more than an aesthetic decision. The
colours need to encode the magnitude of the variable accurately — readers
should be able to intuit that darker means more, or that a colour shift signals
a meaningful boundary. The colours also need to work in print, on screen, and
for readers with colour-vision deficiencies.

mwmap ships 25 named palettes grouped into seven categories. Each palette is a
vector of hex colour codes that has been verified to produce perceptually
distinct and readable maps.

## Browsing the available palettes

`get_mw_palettes()` returns a data frame of all available palettes with their
category, description, and number of colours. This is the quickest way to find
the right palette for your subject matter — filter by `category` to narrow the
list.

```{r}
#| label: tbl-palettes
#| tbl-cap: "All 25 mwmap palettes organised by category."
get_mw_palettes()
```

`view_mw_palettes()` displays all palettes as colour swatches in a ggplot2
chart, which makes it easy to compare options visually before committing to one.

```{r}
#| label: fig-palette-view
#| fig-cap: "All 25 mwmap palettes as colour swatches."
#| fig-height: 9
view_mw_palettes()
```

## Using `scale_fill_mw()` {#sec-scale}

`scale_fill_mw()` is a ggplot2 scale function — drop it into any pipeline
using `+`, just like `scale_fill_distiller()` or `scale_fill_viridis_c()`.

The key arguments:

- `palette` — the name of a palette from `get_mw_palettes()`. Default: `"health"`.
- `discrete` — set `TRUE` for categorical data (factors). Default: `FALSE`
  (continuous gradient).
- `reverse` — `TRUE` flips the palette direction. Useful when the default
  direction (light = low, dark = high) does not suit your variable.
- `name` — the legend title.
- `na.value` — colour for missing data. Defaults to `"grey50"`.

```{r}
#| label: fig-palette-examples
#| fig-cap: "Three palettes applied to the same incidence rate data. The palette choice substantially changes the visual impression."
#| layout-ncol: 3

joined <- mw_join(health_data, quiet = TRUE)

mw_map(data = joined, fill = "rate_per_1000") +
  scale_fill_mw("health", name = "Rate") +
  labs(title = '"health"')

mw_map(data = joined, fill = "rate_per_1000") +
  scale_fill_mw("malaria", name = "Rate") +
  labs(title = '"malaria"')

mw_map(data = joined, fill = "rate_per_1000") +
  scale_fill_mw("poverty", name = "Rate") +
  labs(title = '"poverty"')
```

### Reversing a palette

Some variables make more intuitive sense when the colour direction is flipped.
For example, a "coverage" metric where higher is better might read more
naturally if the dark end of the palette represents high coverage (good) rather
than the default where dark typically implies high risk or high burden.

```{r}
#| label: fig-palette-reverse
#| fig-cap: "The malaria palette reversed. Now lighter colour = higher value."
mw_map(data = joined, fill = "rate_per_1000") +
  scale_fill_mw("malaria", reverse = TRUE, name = "Rate per 1,000") +
  labs(title = "Reversed palette: dark = low, light = high")
```

### Extracting hex codes with `get_mw_palette()`

You can extract the raw hex codes from any palette for use outside of ggplot2 —
in tables, slide decks, custom charts, or base R plots. The `n` argument
interpolates the palette to exactly `n` colours, which is useful when you need
more or fewer colours than the palette's default length.

```{r}
#| label: palette-extract
# All colours in the three-colour 'regions' palette
get_mw_palette("regions")
#> [1] "#E41A1C" "#377EB8" "#4DAF4A"

# Interpolate to exactly 7 colours from the 'health' palette
get_mw_palette("health", n = 7)

# Reverse the malaria palette
get_mw_palette("malaria", reverse = TRUE)
```

---

# Layout and Styling {#sec-layout}

A publication-ready map needs more than data and colour. It needs a descriptive
title, an informative subtitle, a source citation in the caption, and a
well-placed legend. `mw_layout()` is a convenience function that applies all of
these in a single call with consistent typographic styling throughout.

`mw_layout()` returns a special `mw_layout` class object. When you add it to a
ggplot2 map using `+`, it applies both a `labs()` call (for the text elements)
and a `theme()` call (for the visual formatting) — so you get both content and
styling in one step.

```{r}
#| label: fig-layout
#| fig-cap: "The same choropleth polished with mw_layout() for titles, legend placement, and typography."
mw_choropleth(health_data, malaria_cases, palette = "malaria", quiet = TRUE) +
  mw_layout(
    title = "Malaria Cases by District",
    subtitle = "Malawi · Simulated data · 2024",
    caption = "Source: Ministry of Health · Visualisation: mwmap",
    legend_position = "bottom",
    legend_direction = "horizontal",
    legend_title = "Reported Cases"
  )
```

The most useful `mw_layout()` arguments:

| Argument | Description |
|---|---|
| `title`, `subtitle`, `caption` | The three standard ggplot2 text elements |
| `legend_position` | `"right"`, `"left"`, `"bottom"`, `"top"`, or `"none"` |
| `legend_direction` | `"vertical"` (default) or `"horizontal"` — horizontal suits a bottom legend |
| `theme` | Base ggplot2 theme: `"void"` (no axes, default), `"minimal"`, or `"classic"` |
| `panel_border` | `TRUE` adds a thin rectangle around the map panel |
| `background_color` | Plot background fill. Default: `"white"` |
| `grid_major` | `TRUE` adds lat/lon grid lines |
| `font_family` | Font family for all text |
| `title_size` | Title font size. Default: 16 |
| `caption_size` | Caption font size. Default: 9 |
| `legend_key_size` | Size of each legend key in cm |

: `mw_layout()` arguments {tbl-colwidths="[30,70]"}

::: {.callout-tip}
## Legend position for published figures
For figures intended for journal publication or formal reports,
`legend_position = "right"` (the default) is usually the best choice — it
does not compress the map vertically. For slide decks or dashboards where
space is at a premium, `"bottom"` with `legend_direction = "horizontal"`
often works better.
:::

---

# Working with District Names {#sec-names}

District name matching is one of the most common sources of frustration when
working with Malawi data. Administrative names appear differently across
datasets — "Nkhata Bay" vs "Nkhatabay", "Blantyre City" vs "Blantyre", "Mzuzu"
vs "Mzuzu City". The utilities in this section handle these variations
automatically.

## Listing district names with `mw_districts()`

`mw_districts()` returns the official district names as they appear in the
`mw_level_2` shapefile. Use this to build input validation lists, populate
dropdown menus in Shiny apps, or simply check the exact spellings your data
needs to match.

```{r}
#| label: districts-list
# All 28 districts, sorted alphabetically
mw_districts()

# Only Southern region districts ("s" is a valid shortcut for "Southern")
mw_districts("Southern")

# Return a full data frame with standard, admin, and short name variants
# plus the region each district belongs to
mw_districts(type = "all")
```

The `type` argument controls the format of the output:

- `"standard"` (default) — title-case names suitable for display in charts and maps.
- `"admin"` — names exactly as they appear in the shapefile.
- `"short"` — names with suffixes like "City" stripped.
- `"all"` — a data frame with all three variants plus the region column.

## Standardising names with `mw_clean_names()`

`mw_clean_names()` applies a series of transformations to bring district name
strings closer to the shapefile format:

1. Converts to title case (`"lilongwe"` → `"Lilongwe"`)
2. Strips trailing "District", "T/A", "TA" (`"Blantyre District"` → `"Blantyre"`)
3. Normalises whitespace (collapses multiple spaces to one)
4. Applies known shapefile variant mappings:
   - `"Nkhata Bay"` → `"Nkhatabay"` (the shapefile omits the space)
   - `"Mzuzu"` → `"Mzuzu City"` (common shorthand → full shapefile name)

```{r}
#| label: clean-names
messy <- c("lilongwe", "BLANTYRE DISTRICT", "nkhata bay", "mzuzu", "zomba")
mw_clean_names(messy)
#> [1] "Lilongwe"   "Blantyre"   "Nkhatabay"  "Mzuzu City" "Zomba"
```

`mw_join()` calls `mw_clean_names()` on your district column automatically, so
in normal use you never need to call this function yourself. But calling it
explicitly before a join lets you inspect what the cleaning will do to your
data before committing to the join.

## Finding close matches with `mw_suggest_matches()`

When `mw_join()` reports unmatched districts, `mw_suggest_matches()` helps you
figure out the correct name. It uses Levenshtein edit distance to find the most
similar names in the candidate list — the same algorithm used by autocorrect.
The first suggestion is almost always the intended match when you have a simple
typo.

```{r}
#| label: suggest-matches
# A simple typo — one extra "e"
mw_suggest_matches("Lilongwee", mw_districts())
#> [1] "Lilongwe"  "Salima"   "Likoma"

# Another typo
mw_suggest_matches("Blantyree", mw_districts())
#> [1] "Blantyre"  "Balaka"   "Salima"
```

::: {.callout-warning}
## Two name variants to always watch for

| What you might write | Correct shapefile name |
|---|---|
| `"Nkhata Bay"` | `"Nkhatabay"` |
| `"Mzuzu"` | `"Mzuzu City"` |

`mw_clean_names()` and `mw_join()` handle both automatically. But if you are
writing district names directly as strings in `mw_highlight(districts = ...)` or
`mw_labels(districts = ...)`, use the shapefile form (right column).
Check the exact names with `mw_districts()`.
:::

---

# Complete Worked Examples

These examples bring together multiple mwmap functions to produce polished,
multi-layer maps of the kind you might include in a research paper, government
report, or programme dashboard.

## Example 1 — Malaria burden map

This example builds a complete publication-ready map that combines a
categorical choropleth, lake overlay, outline highlights on the major cities,
and text labels. The key insight is that each element is a separate layer added
with `+`, and the final `mw_layout()` call applies consistent titling and
styling to the whole figure.

```{r}
#| label: fig-example-health
#| fig-cap: "A complete, multi-layer malaria burden map suitable for publication."
#| fig-height: 7

# Step 1: classify districts into burden tiers
health_data$burden <- cut(
  health_data$rate_per_1000,
  breaks = c(0, 3, 5, 8, 12, Inf),
  labels = c("Minimal", "Low", "Moderate", "High", "Critical"),
  ordered_result = TRUE
)

# Step 2: join to geometry
map_df <- mw_join(health_data, quiet = TRUE)

# Step 3: build the map layer by layer
mw_map(
  data = map_df,
  fill = "burden",
  lakes = TRUE,
  border_color = "white",
  border_size = 0.4
) +
  scale_fill_mw(
    "malaria",
    discrete = TRUE,
    name = "Burden Level",
    na.value = "grey85"
  ) +
  mw_highlight(
    districts = c("Blantyre", "Lilongwe"),
    outline_only = TRUE,
    color = "#333333",
    size = 1
  ) +
  mw_labels(
    districts = c("Blantyre", "Lilongwe", "Mzuzu City", "Zomba"),
    size = 3,
    color = "#333333"
  ) +
  mw_layout(
    title = "Malaria Burden Classification by District",
    subtitle = "Malawi · District Level · Simulated data · 2024",
    caption = "Source: Simulated data · Visualisation: mwmap package",
    legend_position = "right"
  )
```

---

## Example 2 — Regional vaccination coverage

This example uses `mw_regions()` to show a regional-level indicator. The
darker the green, the higher the vaccination coverage. We encode the coverage
values by manually mapping each region to a colour from the `"vaccination"`
palette — darker for higher coverage. The result communicates both geography
and relative performance at a glance.

```{r}
#| label: fig-example-regions
#| fig-cap: "Vaccination coverage at the regional level. Darker green = higher coverage."
mw_regions(
  region_colors = c(
    "Northern" = "#31A354", # highest coverage → darkest green
    "Central" = "#74C476",
    "Southern" = "#C7E9C0" # lowest coverage → lightest green
  ),
  labels = TRUE,
  label_size = 5,
  label_face = "bold",
  lakes = TRUE
) +
  mw_layout(
    title = "Childhood Vaccination Coverage by Region",
    subtitle = "Malawi · Simulated estimates · 2024",
    caption = "Darker green = higher coverage · Source: Simulated"
  )
```

---

## Example 3 — Infrastructure overlaid on burden data

This example combines two data sources on a single map: the district-level
health burden (continuous choropleth) and the locations of health facilities
(points coloured by facility level). This dual-layer approach allows the reader
to see both the health burden and the distribution of health infrastructure
simultaneously — and to immediately spot areas where burden is high but
facilities are sparse.

```{r}
#| label: fig-example-infra
#| fig-cap: "Health burden choropleth with facility locations overlaid. High-burden areas with few facilities become immediately visible."
#| fig-height: 7

mw_map(
  data = map_df,
  fill = "rate_per_1000",
  lakes = TRUE,
  border_color = "white",
  border_size = 0.3
) +
  scale_fill_mw(
    "health",
    name = "Rate per 1,000",
    na.value = "grey85"
  ) +
  mw_points(
    facilities,
    lon,
    lat,
    color = "level",
    size = 3,
    alpha = 0.9
  ) +
  scale_color_manual(
    values = c("Central" = "#E41A1C", "District" = "#1a5a9a"),
    name = "Facility Level"
  ) +
  mw_layout(
    title = "Health Facilities and Malaria Burden",
    subtitle = "Malawi · 2024 · Simulated data",
    caption = "Blue = district hospitals · Red = central referral hospitals",
    legend_position = "right"
  )
```

---

## Example 4 — Traditional authority detail

This example zooms into a single district (Lilongwe) to show the TA-level
structure. This kind of map is useful for district health offices, local
programme teams, and field research that needs to operate at the sub-district
level. The thick district boundary provides a frame of reference; the TA name
labels let the reader identify each sub-unit by name.

```{r}
#| label: fig-example-ta
#| fig-cap: "Traditional Authorities in Lilongwe district. The thick red border is the district boundary; thin grey lines are TA boundaries."
mw_ta_map(
  districts = "Lilongwe",
  fill = "white",
  border_color = "grey80",
  border_size = 0.2,
  district_border = TRUE,
  district_border_color = "#E41A1C",
  district_border_size = 1.2,
  labels = TRUE,
  label_size = 2.2,
  label_color = "#333333"
) +
  mw_layout(
    title = "Traditional Authorities — Lilongwe District",
    caption = "Source: MASDAP · mwmap package"
  )
```

---

# Tips and Best Practices {#sec-tips}

::: {.callout-tip}
## Always map rates, not raw counts

A map of raw case counts is almost always misleading because it reflects
population size as much as the variable of interest. Lilongwe and Blantyre
will have more of almost everything simply because they are the largest
districts. Before mapping any count variable, divide by the appropriate
denominator:

```r
# Incidence rate per 1,000 population
data$rate_per_1000 <- data$cases / data$population * 1000

# Percentage of target reached
data$coverage_pct <- data$reached / data$target * 100
```

The geographic pattern on a rate map is often dramatically different from a
count map and reveals the underlying epidemiology or social geography in a way
that counts simply cannot.
:::

::: {.callout-tip}
## Build complex maps layer by layer

Rather than trying to pass every option to a single function call, build your
map incrementally and check the output at each step. This makes it far easier
to debug problems — if something looks wrong, print the map after each `+` to
find which layer is causing the issue:

```r
p <- mw_map(data = joined, fill = "rate")
p <- p + scale_fill_mw("malaria", name = "Rate")
p <- p + mw_highlight(c("Lilongwe", "Blantyre"), outline_only = TRUE)
p <- p + mw_labels(districts = c("Lilongwe", "Blantyre"))
p <- p + mw_layout(title = "My Map", caption = "Source: ...")
p
```
:::

::: {.callout-tip}
## Use `quiet = TRUE` in R Markdown and Quarto documents

Several mwmap functions print informational messages to the console by default.
These messages are helpful interactively but clutter rendered documents.
Suppress them with `quiet = TRUE`:

```r
joined <- mw_join(my_data, quiet = TRUE)
mw_choropleth(my_data, my_variable, quiet = TRUE)
```

Alternatively, set `message: false` in the Quarto chunk options to suppress
messages globally for all chunks in the document.
:::

::: {.callout-tip}
## Combine with ggplot2 extension packages

Because `mw_map()` returns a plain ggplot2 object, any ggplot2 extension
package works on top of it without modification:

- **`patchwork`** — arrange multiple maps side by side or in a grid.
- **`ggspatial`** — add scale bars and north arrows (also available via
  `scale_bar = TRUE` and `north_arrow = TRUE` in `mw_map()`).
- **`ggrepel`** — non-overlapping labels via `mw_label_repel()`.
- **`plotly`** — interactive maps via `interactive = TRUE` in `mw_map()` or
  `mw_choropleth()`.
:::

::: {.callout-note}
## Interactive maps with plotly

Pass `interactive = TRUE` to `mw_map()` or `mw_choropleth()` to get a
pannable, zoomable map in HTML documents and Shiny apps:

```r
# Requires: install.packages("plotly")
mw_choropleth(
  health_data,
  rate_per_1000,
  palette     = "malaria",
  title       = "Malaria Incidence Rate",
  interactive = TRUE
)
```

Note that `mw_layout()` and other ggplot2 layer functions cannot be added to a
plotly object with `+`. Apply all ggplot2 modifications before converting to
interactive, or call `plotly::ggplotly(your_ggplot_map)` yourself at the end.
:::

---

# Function Reference Summary {#sec-reference}

For full documentation on any function, including all arguments and additional
examples, use R's built-in help: `?mw_map`, `?mw_choropleth`, `?mw_join`, etc.

| Function | What it does | Key arguments |
|---|---|---|
| `mw_map()` | Main mapping function | `level`, `fill`, `lakes`, `region`, `districts`, `label_districts`, `highlight_districts` |
| `mw_regions()` | Regional overview (three regions) | `fill`, `region_colors`, `labels`, `lakes` |
| `mw_ta_map()` | Traditional Authority map | `districts`, `labels`, `district_border`, `fill_column` |
| `mw_choropleth()` | One-step choropleth | `data`, `value`, `district_col`, `palette`, `scale_type`, `n_breaks` |
| `mw_join()` | Merge a data frame to district polygons | `data`, `district_col`, `map`, `keep_all`, `verbose` |
| `mw_highlight()` | Coloured fill or outline on specific districts | `districts`, `fill`, `color`, `alpha`, `outline_only`, `label` |
| `mw_highlight_multi()` | Different colour per district | Named vector of district → colour pairs |
| `mw_points()` | Add point locations to a map | `data`, `lon`, `lat`, `color`, `size`, `shape`, `label`, `label_column` |
| `mw_labels()` | District name text at centroids | `districts`, `size`, `color`, `fontface`, `label_column` |
| `mw_label_repel()` | Non-overlapping labels (requires ggrepel) | `force`, `max.overlaps` |
| `mw_layout()` | Titles, legend placement, and theme | `title`, `subtitle`, `caption`, `legend_position`, `legend_direction`, `theme`, `panel_border` |
| `scale_fill_mw()` | Apply a named mwmap palette | `palette`, `discrete`, `reverse`, `direction`, `name`, `na.value` |
| `get_mw_palette()` | Extract palette hex codes | `name`, `n`, `reverse`, `interpolate` |
| `get_mw_palettes()` | Data frame of all palettes | `category` |
| `view_mw_palettes()` | Swatch chart of all palettes | `n`, `ncol` |
| `mw_districts()` | District names as a vector or data frame | `region`, `type`, `sorted`, `include_ta` |
| `mw_districts_simple()` | Simple character vector of district names | `region` |
| `mw_clean_names()` | Standardise district name strings | `x` |
| `mw_suggest_matches()` | Find closest shapefile names by edit distance | `x`, `candidates`, `n` |

: Complete mwmap function reference {tbl-colwidths="[24,41,35]"}

---

# Session Information {.appendix}

The output below records the R version and package versions used when this
document was produced. This information is useful for reproducing results and
diagnosing version-related issues.

```{r}
#| label: session-info
sessionInfo()
```

---

::: {.callout-note appearance="minimal"}
**mwmap** is developed and maintained by Chifundo Bita.
Source code and issue tracker: [github.com/bitacanalytics/mwmap](https://github.com/bitacanalytics/mwmap)
Install from CRAN: `install.packages("mwmap")`
Install development version: `remotes::install_github("bitacanalytics/mwmap")`
:::
